Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique

### 2017-11-06 inst/rmd/cop_2017.Rmd
presentation update

---
title: "Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique"
subtitle: 'IHMT, October, 2017, Lisboa, Portugal economicsofmalaria.com/ihmt'
fig_height: 5
fig_width: 7
bibliography: bibliography.bib
output:
  beamer_presentation:
    includes:
      in_header: preamble_beamer.sty
    theme: "Berlin"
    colortheme: "seagull"
    fonttheme: "structuresmallcapsserif"
---

@@

# Define function fro breaking the days_since var
break_days_since <- function(days_since){
  out <- ifelse(days_since > 270, '270+',
    ifelse(days_since > 180, '181-270',
           ifelse(days_since > 90, '091-180',
                  ifelse(days_since > 60, '061-090',
                         ifelse(days_since > 30, '031-060',
                                ifelse(days_since >= 0, '000-030',
                                       ifelse(days_since < 0, 'Before', NA)))))))
  out <- factor(out,
                levels = c('Before',
                           '000-030',
                           '031-060',
                           '061-090',
                           '091-180',
                           '181-270',
                           '270+'))
  return(out)
}

@@

# Create model data for all obs (not just propensity score)
model_data <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs) %>%
  # left_join(matched) %>%
  mutate(days_since = break_days_since(days_since)) %>%
  filter(!is.na(days_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick)) %>%
  mutate(month = format(date, '%b'))
  # # remove the period 30 days after spraying
  # filter(days_since >= 30 | days_since <= 0)
# Need to add seasonality
# Need to adjust for houses on and off site

@@

@@

- Malaria has a nearly unquantifiably large economic impact.
- Many channels: fertility, fecundity, saving, investment [@Shretta2016], risk perception, productivity, absenteeism, human capital accumulation [@CastelBranco2014], mortality, costs of care [@Sachs2002].
- Cost-benefit studies often only consider the costs of an intervention and associated costs of care, without quantifying the societal cost of non-intervention.

@@

## Regression discontunity analysis

@@

@@

- 30-50% reduction in absenteeism in the 6 months after IRS, when administered during malaria season. 
- Next steps are incorporation of (a) productivity data (via cane cut), (b) better clinical data (via local health facilities), and (c) full cost data.  
- Will also be comparing with a sugarcane facility in a zone where an elimination campaign is taking place.

## Implications

- Reduction is large and significant.
- Depending on detailed cost data (pending), IRS may be effective even from a purely financial point of view (ie, beyong just "corporate social responsibility").

@@

- No analysis yet of different worker types (agricultural vs. industrial).
- Have not yet ventured at all into side-analyses (effect on employment, tonnage, etc.). 
- Sick absenteeism seems to track absenteeism poorly: lack of clarity regarding pathways.

@@

**Presentation:** [economicsofmalaria.com/ihmt](http://economicsofmalaria.com/ihmt)

>>>>

---
title: "Effectiveness of private sector malaria control: The case of sugarcane workers in Mozambique"
subtitle: 'Malaria Economics Community of Practice, November, 2017, Baltimore, Maryland, USA economicsofmalaria.com/cop.pdf'
fig_height: 5
fig_width: 7
bibliography: bibliography.bib
output:
  beamer_presentation:
    includes:
      in_header: preamble_beamer.sty
    theme: "Berlin"
    colortheme: "seagull"
    fonttheme: "structuresmallcapsserif"
---

@@

# Define function fro breaking the days_since var
break_days_since <- function(days_since){
  out <- ifelse(days_since > 180, '181+',
           ifelse(days_since > 90, '091-180',
                  ifelse(days_since > 60, '061-090',
                         ifelse(days_since > 30, '031-060',
                                ifelse(days_since >= 0, '000-030',
                                       ifelse(days_since < 0, 'Before', NA))))))
  out <- factor(out,
                levels = c('Before',
                           '000-030',
                           '031-060',
                           '061-090',
                           '091-180',
                           '181+'))
  return(out)
}

@@

# Create model data for all obs (not just propensity score)
model_data <-
  ab_panel %>%
  # filter(oracle_number %in% matched$oracle_number) %>%
  left_join(irs) %>%
  # left_join(matched) %>%
  mutate(spray_month = format(date + days_since, '%b')) %>%
  mutate(days_since = break_days_since(days_since)) %>%
  filter(!is.na(days_since)) %>%
  # mutate(time_period = lendable::time_period_extract(date)) %>%
  # mutate(time_period = as.character(time_period)) %>%
  mutate(time_period = make_season(date = date)) %>%
  mutate(absent_sick = ifelse(is.na(absent_sick), FALSE, absent_sick)) %>%
  mutate(month = format(date, '%b'))
  # # remove the period 30 days after spraying
  # filter(days_since >= 30 | days_since <= 0)
# Need to add seasonality
# Need to adjust for houses on and off site

@@

# new fit
# new_fit <- lm(absent ~ days_since * time_period + month, data = model_data)

@@

- Malaria has a nearly unquantifiably large economic impact.
- Many channels: fertility, fecundity, saving, investment [@Shretta2016], risk perception, productivity, absenteeism, human capital accumulation [@CastelBranco2014], mortality, costs of care [@Sachs2002].
- Cost-benefit studies often only consider the costs of an intervention and associated costs of care, without quantifying the societal cost of non-intervention.  
- In elimination context, scaling-up private sector involvement is very appealing.

@@

## Regression discontinuity analysis

@@

## Back of the envelope calculations

```{r}
# Get yearly absences / presences
x <- model_data %>%
  filter(time_period == 'Malaria season') %>%
  group_by(days_since) %>%
  summarise(absences = length(which(absent)),
            eligibles = n()) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100)
```

**Savings**

- In percentage point terms, reduction from 13% to 8%.  
- 5 annually prevented absences per worker.   
- 8,000 workers: 40,000 prevented absences, wage of 3 USD
- TOTAL: 120,000 USD in productivity-only savings.

**Costs**

- 8 IRS workers, 1500 USD yearly = 12,000 USD  
- ACT + DDT: 50,000 USD 
- Facilities, vehicles, gas: 50,000 USD  
- TOTAL: 112,000 USD in IRS-only costs

**7% ROI** (ignoring clinical costs)

@@

- 30-50% reduction in absenteeism in the 6 months after IRS during malaria season. 
- Depending on detailed cost data (pending), IRS may be effective even from a purely financial point of view (ie, beyond just "corporate social responsibility").  
- Next steps are incorporation of (a) productivity data (via cane cut), (b) better clinical data (via local health facilities), and (c) full cost data.  
- Will also be comparing with a sugarcane facility in a zone where an elimination campaign is taking place.

@@

- No analysis yet of different worker types (agricultural vs. industrial).
- Have not yet ventured at all into side-analyses (effect on employment, tonnage, etc.). 
- Sick absenteeism seems to track absenteeism poorly: lack of clarity regarding pathways.  
- Large sample size, but all from same place: questionable generalizability.

@@

**Presentation:** [economicsofmalaria.com/ihmt.pdf](http://economicsofmalaria.com/ihmt.pdf)
